---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
    In order to load and preprocess the data, we use the dplyr and lubridate
    libraries, and then we group the data by date, and summarize the data to
    calculate the daily totals.
``` {r message=FALSE}
library(dplyr)
library(lubridate)
```
```{r}

df<-read.csv("activity.csv", stringsAsFactors = FALSE)
df$date<-as.Date(as.character(df$date))
by_date<-group_by(df,date)
sum_by_date<-summarise(by_date, daily_total=sum(steps,na.rm=TRUE))

```


## What is mean total number of steps taken per day?
    The mean and medians for the data are calculated on the grouped data 
    and the total sums are plotted using a barplot
```{r}

# Total number of steps divided by number of dates
meanByDate<-sum(sum_by_date$daily_total,na.rm=TRUE)/length(sum_by_date$date)
median_step<-median(sum_by_date$daily_total, na.rm=TRUE)
barplot(daily_total ~ date, 
        data=sum_by_date,
        xlab="Date",
        ylab="Total Daily Steps",
        main="Total Steps per Day")


```
    
    The averages for the data are below:
      * The mean number of steps per days is `r meanByDate`
      * The median number of steps per day is `r median_step`

## What is the average daily activity pattern?

    In order to calculate the average daily pattern, we must first group by the
    interval, and then summarize the data.  The results are plotted in a line plot
    below.
```{r}

by_interval<-group_by(df,interval)
interval_summary<-summarize(by_interval, interval_sum=sum(steps,na.rm=TRUE),
                           interval_mean=mean(steps,na.rm=TRUE),
                           interval_median=median(steps,na.rm=TRUE))

with(interval_summary, plot(x=interval, y=interval_mean, 
                            type="l",
                            xlab="Interval", 
                            ylab="Mean number of steps",
                            main="Mean Number of Steps by Interval"))
```
    
    Even though the assignment required a line plot, a bar plot seems to provide
    better interval resolution. We can use this plot to further verify the 
    calculation of the interval with the maximum number of steps.
```{r}
max_mean_steps<-max(interval_summary$interval_mean)
max_row<-filter(interval_summary, interval_mean >= max_mean_steps)
barplot(interval_mean ~ interval,data=interval_summary,
        xlab="Interval", 
        ylab="Mean number of steps",
        main="Mean Number of Steps by Interval")


```
  
  * The interval containing the maximum mean steps is `r max_row$interval`

## Imputing missing values

    For this dataset, we will impute the data using the mean steps for the 
    day to replace NA values. Once the data is imputed, it can be summarized
    and plotted with a barplot to show the differences with the non-imputed data.

``` {r message=FALSE}
library(Hmisc)
```
``` {r}
#Calculate missing values
num_missing<-sum(is.na(df$steps))
#impute the data from the original dataframe
df$imputed_steps<-with(df, impute(steps,mean))
fixed_data<-select(df, c(date, interval,imputed_steps))

#Group and Summarize the imputed data
by_date<-group_by(df,date)
imputed_sum<-summarise(by_date,daily_total=sum(imputed_steps), 
                              daily_mean=mean(imputed_steps),
                              daily_median=median(imputed_steps))
imputed_median<-median(imputed_sum$daily_total)
imputed_mean<-sum(imputed_sum$daily_total)/length(imputed_sum$date)
barplot(daily_total ~ date, 
        data=imputed_sum,
        xlab="Date",
        ylab="Total Daily Steps",
        main="Total Steps per Day")
```
  
  * The number of missing values in the data is `r num_missing`
  
  * After imputing the data: 
      * the mean number of steps per day is now `r format(imputed_mean, digits=8)`
      * the median number of steps per day is now `r format(imputed_median, digits=8)`
  
## Are there differences in activity patterns between weekdays and weekends?

    In order to compare the weekday vs weekend activity, a factor variable must 
    first be created that distinguishes between week day vs weekend data. 
    Once the data is filtered on those two factors, it can be grouped, summarized
    and plotted to show differences in activity patterns.
``` {r}

#create a factor variable for the imputed data (weekend vs weekday)
fixed_data<-mutate(fixed_data, 
                   day_of_week=factor(ifelse(weekdays(date, abbreviate=TRUE) %in%
                                        c("Sat","Sun"),"weekend","weekday")))
#group the data by interval
fixed_by_interval<-group_by(fixed_data,interval)

#filter the weekday data and summarize
weekday_data<-filter(fixed_by_interval, day_of_week=="weekday")
summarized_weekday<-summarise(weekday_data, interval_mean=mean(imputed_steps),interval,date)

#filter the weekend data and summarize
weekend_data<-filter(fixed_by_interval, day_of_week=="weekend")
summarized_weekend<-summarise(weekend_data, interval_mean=mean(imputed_steps),interval,date)

#Make side by side plots of the weekend and weekday data
par(mfrow=c(2,1))
with(summarized_weekday, plot(x=interval, y=interval_mean, 
                            type="l",
                            xlab="Interval",
                            ylab="Mean number of steps",
                            main="Mean Number of Steps by Interval Weekday"))
with(summarized_weekend, plot(x=interval, y=interval_mean, 
                            type="l",
                            xlab="Interval",
                            ylab="Mean number of steps",
                            main="Mean Number of Steps by Interval Weekend"))


```
